Equivalence Hypergraphs: E-Graphs for Monoidal Theories

### 2024-01-08 E-Hypergraphs-LICS24.tex
fix

This normal form can be used to prove the following completeness theorem. 
\begin{theorem}[Completeness]
We have the following equivalence of categories
\[
	\textbf{SMT}^{+}(\Sigma, \mathcal {E} ) \cong \WellTypedMdaEcospans / \mathcal{S,E}~.
\]
\end{theorem}
\begin{proof}
The equivalence is given by the functor $\llbracket - \rrbracket$.  First, note that the functor is surjective on objects,  so it suffices to prove the functor is both full and faithful.  In each case,  the result follows by a normal form argument in combination with relying on the equivalence $(*)$ $\textbf{SMT}(\Sigma, \mathcal E) \cong \textbf{MHypI}(\Sigma)/\mathcal{E}$ of symmetric monoidal theories and (appropriately quotiented) plain MDA hypergraphs which is given by the restriction of $\llbracket - \rrbracket$ to $$\textbf{SMT}(\Sigma, \mathcal E)$ which is a subcategory of $\textbf{SMT}^{+}(\Sigma, \mathcal {E} )$.

To see that the functor is faithful,  consider $f$ and $g$ of $\textbf{SMT}^{+}(\Sigma, \mathcal {E} )$ such that $\llbracket f \rrbracket = \llrbracket g \rrbracket$.  We aim to show that $f = g$.  First,  find weak normal forms of $\llbracket f \rrbracket$ and $\llrbracket g \rrbracket$,  namely $\phi_1 + \ldots + \phi_ni$ and $\psi_1 + \ldots + \psi_n$.  We now need a further observation that if $\llbracket f \rrbracket \Rrightarrow_{{\llbracket \mathcal{S,E} \rrbracket}^+} \llrbracket g \rrbracket$,  then there exist rewrites $\phi_i \Rrightarrow_{\mathcal E} \psi_i$.  Thus,  by $(*)$ we have that $\llbracket \phi_i \rrbracket^{-1} = \llbracket \psi_i \rrbracket^{-1}$.  Noting that $f = \llbracket \phi_1 \rrbracket^{-1} + \ldots + \llbracket \phi_n \rrbracket^{-1}$ and $g = \llbracket \psi_1 \rrbracket^{-1} + \ldots + \llrbacket \psi_n \rrbracket^{-1}. 
CHRIS: THIS NEEDS FIXING.

To see that the functor is full,  consider a cospan $f$ in $\WellTypedMdaEcospans$.  We can find an equivalent weak normal form $f_1 + \ldots + f_n$,  where each $f_i$ is a plain MDA hypergraph.  Thus,  by $(*)$ we can find morphisms $g_i$ such that $\llbracket g_i \rrbracket = f_i$.  Thus,  we have a morphism $g_1 + \ldots + g_n$ such that $\llbracket g_1 + \ldots + g_n \rrbracket = \llbracket g_1 \rrbracket + \ldots \llbracket g_n \rrbracket = f_1 + \ldots + f_n$,  as required. 
\end{proof}

>>>>

This normal form can be used to prove the following completeness theorem. 
\begin{theorem}[Completeness]
We have the following equivalence of categories
\[
	\textbf{SMT}^{+}(\Sigma, \mathcal {E} ) \cong \WellTypedMdaEcospans / \mathcal{S,E}~.
\]
\end{theorem}
\begin{proof}
The equivalence is given by the functor $\llbracket - \rrbracket$.  First, note that the functor is surjective on objects,  so it suffices to prove the functor is both full and faithful.  In each case,  the result follows by a normal form argument in combination with relying on the equivalence $\textbf{SMT}(\Sigma, \mathcal E) \cong \textbf{MHypI}(\Sigma)/\mathcal{E}$ of symmetric monoidal theories and (appropriately quotiented) plain MDA hypergraphs which is given by the restriction of $\llbracket - \rrbracket$ to $\textbf{SMT}(\Sigma, \mathcal E)$ which is a subcategory of $\textbf{SMT}^{+}(\Sigma, \mathcal {E} )$.

%To see that the functor is faithful,  consider $f$ and $g$ of $\textbf{SMT}^{+}(\Sigma, \mathcal {E} )$ such that $\llbracket f \rrbracket = \llbracket g \rrbracket$.  We aim to show that $f = g$.  First,  find weak normal forms of $\llbracket f \rrbracket$ and $\llbracket g \rrbracket$,  namely $\phi_1 + \ldots + \phi_ni$ and $\psi_1 + \ldots + \psi_n$.  We now need a further observation that if $\llbracket f \rrbracket \Rrightarrow_{{\llbracket \mathcal{S,E} \rrbracket}^+} \llbracket g \rrbracket$,  then there exist rewrites $\phi_i \Rrightarrow_{\mathcal E} \psi_i$.  Thus,  by $(*)$ we have that $\llbracket \phi_i \rrbracket^{-1} = \llbracket \psi_i \rrbracket^{-1}$.  Noting that $f = \llbracket \phi_1 \rrbracket^{-1} + \ldots + \llbracket \phi_n \rrbracket^{-1}$ and $g = \llbracket \psi_1 \rrbracket^{-1} + \ldots + \llrbacket \psi_n \rrbracket^{-1}. 
%CHRIS: THIS NEEDS FIXING. 
%
To see that the functor is full,  consider a cospan $f$ in $\WellTypedMdaEcospans$.  We can find an equivalent weak normal form $f_1 + \ldots + f_n$,  where each $f_i$ is a plain MDA hypergraph.  Thus,  by $(*)$ we can find morphisms $g_i$ such that $\llbracket g_i \rrbracket = f_i$.  Thus,  we have a morphism $g_1 + \ldots + g_n$ such that $\llbracket g_1 + \ldots + g_n \rrbracket = \llbracket g_1 \rrbracket + \ldots \llbracket g_n \rrbracket = f_1 + \ldots + f_n$,  as required. 
\end{proof}
